Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests

### 2018-11-20 review/response.md

2 Reviewer 1
------------

2.1 Overall Summary
-------------------

This paper is about the estimation of pedestrian footfall using Wi-Fi
signal emitted by mobile devices.
It deals with the problem of estimating the footfall size based on a noisy
and biased data. More specifically it addresses the following issues that may
lead to overcounting:
 - MAC randomization that may cause a device to generate several MAC addresses;
 - Signals received from outside the interest zone; unequal adoption
   of Wi-Fi smartphone in the population.
To address this issue, a number of techniques are introduced:
 - clustering received signal strength to remove device that are outside
   the monitored area;
 - clustering probes based on sequence number and timing;
 - finally an adjustment factor is applied to correct the estimated value.
   The adjustment factor is calibrated for each location using
   a manual counting.
The author conducted a pilot experiment to identify the clustering 
algorithms and parameters, and then they conducted a second range of 
experiments at different location over an extended period of time. During 
each of those experiments, ground-truth count was collected manually.
The results show that the presented approach can significantly reduce the
counting error compared to a naive solution consisting in counting the 
distinct number of MAC observed.
Nevertheless, there are cases (2) where the error is still large (50%),
even if in other cases (1) the error is only of 9%.

> This is a correct surmisal of the remit of the paper. We have also added
> material to the introduction and conclusion of the paper to better link
> this remit to the Smart Cities discourse.

@@

> Vicente R, Wibral M, Lindner M, Pipa G. Transfer entropy--a model-free 
> measure of effective connectivity for the neurosciences.
> J Comput Neurosci. 2010;30(1):45-67.

> we can infer the relation or flow between the locations using other
> probabilistic approaches.

7. P10 l48, how was the adjustment factor determined?

> The adjustment factor is calculated as the ratio between the manual counts to
> the counts calculated from sensor for the sample period. This 
> is used to adjust future counts at the same locations. We have included
> the clarification on the revised manuscript.

@@

> Lugomer, K., & Longley, P. (2018). Towards a Comprehensive Temporal 
> Classification of Footfall Patterns in the Cities of Great Britain (Short
> Paper). In LIPIcs-Leibniz International Proceedings in Informatics (Vol. 114).
> Schloss Dagstuhl-Leibniz-Zentrum fuer Informatik.

> There was no reason for
> selecting the time in the pilot study. the location was chosen for it it known
> to have high number of randomised probe requests and has one of the highest
> footfall in London. For the main study the counting was done such that each
> location has atleast 2/3 counts in the same time interval.

2. Will you give more detailed information about the data and the parameters for 
various clustering algorithms? Including the number of records, the number of 
clusters. How about the performance for different clustering methods 
in your study？

> Thank you. This part has been expanded and clarified in the revised manuscript. 
> The performance of the algorithms used for one dimensional classification were 
> evaluated on the basis of the time taken for processing and the resulting 
> mean absolute percentage error. We found that the k-means gives
> us the best data for the quickest processing time. The graph based clustering
> was done using two parameters - sequence threshold of 60 and time threshold of
> 16 seconds. These values were arrived arrived at using trial and error by using
> a know device as the control mechanism. This has been illustrated further in 
> the revised manuscript.

@@

3. Methodology: Will you draw a diagram of the Methodology?

> We have resequenced some of the material in response to Referee 2
> and, cognisant of over-all length requirements, think that this (rather
> than a diagram) addresses your concerns.

4. Performance Analysis: How about the performance for collecting data using 
PostgreSQL? Did you compare Apache Flink and Spark Streaming for real-time data 
collecting?

> This is an interesting point in the context of big data and tools for tackling
> big data. Though the data set here is large enough to make desktop computing
> unfeasible, it is neither big nor fast enough for a large scale big data
> system. An opensource and free toolkit comprising of text processing tools 
> and relational database management system was enough 
> to handle the compexity, size and speed of the data.
> The toolkit employed here involves PostgreSQL, Bash, Unix tools
> for text processing (sed, awk, sort, join), Node.js and R.
> The same toolkit proved to be useful in the larger project 
> (smart street sensor) providing near real-time delivery of data.

### 2018-11-20 paper/sections/conclusion.tex
Tue 20 Nov 13:38:53 GMT 2018

This paper has made a novel contribution to the measurement of activities in retail centres in real time.
To this end, it has described the collection and processing of a novel consumer Big Dataset that enables valid measures of levels of footfall activity which has been scaled across a Great Britain wide network of sensors.
In both conceptual and technical terms, it illustrates the ways in which passively collected consumer data can be ‘hardened’ to render them robust and reliable by using related procedures of internal and external validation.

Internal validation addresses the issues of screening out device probes that do not indicate footfall, and the further screening of device probes to ‘fingerprint’ the effects of MAC randomization.It is important to note that the filtering process work based soley on the information present in the probe requests and their temporal distribution.
This ensures that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves. 
External validation then entails reconciling adjusted counts with the footfall observed at sample locations.
This procedure makes it possible to generalise from locations at which manual footfall surveys are conducted to all others in the system, and to develop a classification of device locations that are more or less susceptible to noise generation.

This Wi-Fi based footfall counting methodology offers a large number of applications and benefits for real time spatial analysis.
Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
A snapshot showing week's worth of precise footfall in area around Charring cross, London is shown in Figure \ref{} to demonstrate the potential for such dataset.
Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.
Such precise and granular data also enables us to confidently model the pedestrian flow in urban road networks, and will be an indispensable tool in the smart city framework.
It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them which we intend to research in future.

>>>>

Sentient technologies make possible measurement of the human activities that are
the life blood of the smart city. Yet the data that they harvest are frequently
relevant only to sub-groups within society that avail themselves of particular
goods and services – such as social media applications, transport modes or
retail offers. In each of these cases, it is necessary to remember that the
resulting data are a by-products of consumer transactions, and will as a
consequence, only pertain to users of the relevant good or service. If the smart
city is to be socially inclusive, it therefore follows that sentient data must
represent entire populations, whether by design or by triangulation with
external, population wide, sources. This usually poses a non-trivial task, since
the ebbs and flows of smart device-enabled citizens rarely pertain to any
clearly defined population in either administrative or functional terms Massam
1973.

Our objective here has been to collect, rather than re-use, data on smart city
functioning, by recording Wi-Fi probes and ultimately reconciling them with
manual counts in order to infer ambient populations. The internal validation
methodology set out in the technical sections of this paper, allied to external
validation from pedestrian counts, renders the method inclusive and robust when
recording activity levels in retail centres in real time. We have described the
collection and processing of a novel consumer Big Dataset that enables valid
measures of levels of footfall activity which has been scaled across a Great
Britain wide network of sensors CDRC 2016.  In both conceptual and technical
terms, it illustrates the ways in which passively collected consumer data can be
‘hardened’ to render them robust and reliable by using related procedures of
internal and external validation.

Internal validation addresses the issues of screening out device probes that do
not indicate footfall, and the further screening of device probes to
‘fingerprint’ the effects of MAC randomization. It is important to note that the
filtering process work based solely on the information present in the probe
requests and their temporal distribution. This ensures that although the mobile
devices were uniquely identified, there was no further personal data generated
by linking the probe requests to the users of the mobile devices. This method
essentially gave us a way to estimate the footfall in real-time without
identifying or tracking the mobile devices themselves. External validation then
entails reconciling adjusted counts with the footfall observed at sample
locations. This procedure makes it possible to generalise from locations at
which manual footfall surveys are conducted to all others in the system, and to
develop a classification of device locations that are more or less susceptible
to noise generation.

This Wi-Fi based footfall counting methodology offers a large number of
applications and benefits for real time spatial analysis. Since Wi-Fi based
sensors are inexpensive and the data model is scalable, it is possible to use
this methodology for a large network of sensors to gather granular data on
pedestrian footfall. A snapshot showing week’s worth of precise footfall in area
around Charring cross, London is shown in Figure 7 to demonstrate the potential
for such dataset. Projects such as SmartStreetSensors CDRC 2016, may utilise
this methodology to overcome the challenges introduced by the implementation of
MAC address randomisation.

The vicissitudes of MAC randomisation, and the provisions of privacy legislation
such as EU General Data Protection Regulations mitigate against tracking
individuals across the smart city using this approach.  This can be modelled
using agent-based methods Heppenstall et al 2012, however. In our own research
we have also begun to link store time-lagged till receipts to footfall, and have
used such data to better understand the dwell times that characterise such
different retail uses as stores with window displays and fast food restaurants.
Such analysis not only provides a more nuanced picture of movement through
retail areas, but also enables valorisation of micro sites within retail
centres. In the UK, for example, this is of immediate practical importance in
evaluating business rates on properties, and has still wider implications for
the setting of retail unit rental values. There are obvious extensions to
understanding the ebbs and flows of activities in the 24 hour smart city.

More broadly still, extensions to this strand of smart city research are likely
to seek to differentiate the quality of different elements within footfall
according to mission e.g. travel to adjacent workplace zones, leisure, etc and
personal characteristics such as spending power. In this respect, future
research may not only simulate linkage of harmonised footfall counts between
sensor locations, but also link these in turn to disaggregate origin-destination
matrices for bikeshare and other public transport modes.  Our own investigations
will consider these and other challenges to understanding the functioning of the
sentient city.

### 2018-11-20 paper/sections/introduction.tex
Tue 20 Nov 13:38:53 GMT 2018

New and developing technologies today provide the infrastructure over which movements and interactions of people can be measured and monitored in the `sentient city' \citep{amin2017seeing}.
These include mobile phone networks, which can triangulate user locations relative to networks of masts, use of GPS to locate users of social media services, and Wi-Fi beacon connectivity to access the Internet.
These technologies offer differing levels of spatial precision, where mobile telephony and Wi-Fi generally being less reliable and offer lower precision than GPS to the end users, while being more advantageous for broader mobility studies \citep{pinelli2015comparing}.

There has been considerable research into the usefulness of these technologies to understand patterns of movement in cities in near real time \citep{candia2008uncovering, gonzalez2008understanding, calabrese2013understanding}. Most of this research has focused upon technical specification of accuracy or precision \citep{song2010limits, lane2010survey}, with somewhat less attention devoted to the ways that the characteristics of the technologies and of their human users conspire to create possible bias in the inclusiveness of usage patterns.
This is perhaps most manifest in the plethora of papers that analyse the distribution and content of publicly available Twitter social media feeds, which likely constitute an extremely biased and self-selecting subset of users that geo-enable their posts, drawn from a user base that itself bears no identified correspondence to any known background population \citep{lansley2016deriving}. 
Even analysis of mobile phone data, usually derived from industry players that have significant market share and user bases representative of local populations, may exclude groups such as tourists from distant origins or subscribers to third party services that share distinctive characteristics \citep{di2016mind}

@@

In this paper, through a set of experiments, we evaluate the value of data collected from a Great Britain wide network of 800 devices \citep{sss2016}, installed in order to characterize the footfall patterns of a scientifically balanced sample of retail centres.
These devices are located in shop windows and record the probes emitted by mobile phones and other Wi-Fi enabled devices.
The data collected from these devices are deemed to be consumer data because devices carried by consumers routinely probe for Wi-Fi connection which is a consumer service.
Monitoring the probes from such devices provides an indication of the presence of their users, regardless of whether or not internet connectivity is established.
Our core motivation is to appraise the usefulness of Wi-Fi probe requests harvested from our network of sensors in order to indicate levels of pedestrian activity.
More broadly still, in our future research we intend to classify the nationwide network of footfall profiles as part of a programme of research to understand the form and functioning of retail areas at a time of far-reaching structural change for the retail industry.

To this end, it is important to first undertake a thorough conceptual and technical appraisal of our consumer data source.
In technical terms, screening the information present in the `probes requests' and classifying them based on their characteristics is essential in order to remove the ones emitted by devices that do not indicate pedestrian activity, such as network enabled printers and other fixed devices.
Related to this, a method to fingerprint Wi-Fi probes is necessary to remove probes from individuals’ devices that in conceptual terms should not be considered part of footfall – as when, for example, an employees is seated in an office within range of the sensor device.
A calibration of sensor measurement is also essential on two grounds: first, individuals may carry multiple devices, or no device at all; and second, the positioning and orientation of the sensor in the retail unit may lead to systematic over- or under-enumeration.
These sources of bias in measurement must be accommodated by manual recording of footfall at each location and generalization of these sample survey results to all locations and time periods.
As we describe in detail below, manual validation of the data needs to be undertaken in parallel with technical profiling of the mix of consumer mobile devices that probe our sensors, since the effectiveness of data cleaning procedures discussed in this paper differ between individual locations and configurations.

>>>>

New and developing ‘smart’ technologies today provide the infrastructure over
which movements and interactions of people can be measured and monitored in the
‘sentient city’ (Amin and Thrift 2017). This is making it possible to
reinvigorate conceptualise the city as the locus of human activities
supplementing night time geographies of residence (Martin et al 2015) with
geographies of shopping behaviour (Lloyd and Cheshire 2018*), workzone
geographies (Singleton and Longley 2018) and studies of movement trajectories
(Campbell et al 2008). This is rendering activity-based conceptions of human
behaviour central to analysis of hardship and opportunity in and around the
smart city (Venerandi et al 2015).

Sentient technologies include mobile phone networks, which can triangulate user
locations relative to networks of masts, use of GPS to locate users of social
media services, and Wi-Fi beacon connectivity to access the Internet. These
technologies offer differing levels of spatial precision, where mobile telephony
and Wi-Fi generally being less reliable and offer lower precision than GPS to
the end users, while being more advantageous for broader mobility studies
(Pinelli et al.  2015). There has been considerable research into the usefulness
of these technologies to understand patterns of movement in cities in near real
time (Candia et al. 2008, Gonzalez et al. 2008, Calabrese et al. 2013).  Most of
this research has focused upon technical specification of accuracy or precision
(Song et al. 2010, Lane et al. 2010), with somewhat less attention devoted to
the ways that the characteristics of the technologies and of their human users
conspire to create possible bias in representing usage patterns across the
entire smart city.

Even analysis of mobile phone data, usually derived from industry players that
have significant market share and user bases representative of local
populations, may exclude groups such as tourists from distant origins or
subscribers to third party services that share distinctive characteristics
\citep{di2016mind}

@@

In this paper, through a set of experiments, we evaluate the value of data
collected from a Great Britain wide network of 800 devices \citep{sss2016}
installed in order to characterize the footfall patterns of a scientifically
balanced sample of retail centres.  These devices are located in shop windows
and record the probes emitted by mobile phones and other Wi-Fi enabled devices.
The data collected from these devices are deemed to be consumer data because
devices carried by consumers routinely probe for Wi-Fi connection which is a
consumer service.  Monitoring the probes from such devices provides an
indication of the presence of their users, regardless of whether or not internet
connectivity is established.  Our core motivation is to appraise the usefulness
of Wi-Fi probe requests harvested from our network of sensors in order to
indicate levels of pedestrian activity.  More broadly still, in our future
research we intend to classify the nationwide network of footfall profiles as
part of a programme of research to understand the form and functioning of retail
areas at a time of far-reaching structural change for the retail industry.

To this end, it is important to first undertake a thorough conceptual and
technical appraisal of our consumer data source.  In technical terms, screening
the information present in the `probes requests' and classifying them based on
their characteristics is essential in order to remove the ones emitted by
devices that do not indicate pedestrian activity, such as network enabled
printers and other fixed devices.  Related to this, a method to fingerprint
Wi-Fi probes is necessary to remove probes from individuals’ devices that in
conceptual terms should not be considered part of footfall – as when, for
example, an employees is seated in an office within range of the sensor device.
A calibration of sensor measurement is also essential on two grounds: first
individuals may carry multiple devices, or no device at all; and second, the
positioning and orientation of the sensor in the retail unit may lead to
systematic over- or under-enumeration.  These sources of bias in measurement
must be accommodated by manual recording of footfall at each location and
generalization of these sample survey results to all locations and time periods.
As we describe in detail below, manual validation of the data needs to be
undertaken in parallel with technical profiling of the mix of consumer mobile
devices that probe our sensors, since the effectiveness of data cleaning
procedures discussed in this paper differ between individual locations and
configurations.

### 2018-11-20 paper/sections/main_study.tex
Tue 20 Nov 13:38:53 GMT 2018

The main study was designed to validate the results of the pilot study over different locations at different times. We start by choosing five different locations across central London, install the sensors and collect data for extended periods of time. 
We also carry out manual counting at these locations across different times of the day. 
We then apply the methodologies discussed earlier to arrive at estimated pedestrian footfall and compare them with corresponding manual counts.
We finally evaluate the effectiveness of the processes with the Mean Absolute Percentage Error (MAPE) at these locations and report our findings.

@@

We start by looking at the distribution of the signal strength reported by the probe requests across the locations. From the density plot shown in Figure \ref{main_signal_strength} we can see that there is clear relation between the distribution of the signal strength and the distance and complexity of the source of noise. We can see that while location 5 shows clean difference between low and high signal strengths, location 2 is almost normally distributed.
Intuitively we expect that location 2 and 4 must be harder to classify than location 1 3 and 5. 
We run the k-means classification algorithm and filter out the probe requests which are randomised and have signal strengths less than the second break (threshold). 
It is important to note that we are dealing with relative thresholds of signal strengths which can vary with location and time of the analysis.
We then aggregate then probe requests for every minute by counting the number of Unique MAC addresses present in every minute. We also remove devices that dwell around the sensor by removing the MAC addresses which reappear from within the past hour.
The results of the first stage of filtering process along with the thresholds are shown in Table \ref{errors_table}. Confirming our intuition, we see that the location 2 has the most MAPE followed by location 4 while rest of them have highly reduced MAPE.
It is significant that this method alone reduces our margin of error by a 50 - 100\% from the raw counts without any cleaning.
This makes the signal strength filtering quick and ideal method for practical applications which doesn't require absolute numbers such as creating large aggregated indexes to show long term trends. 
We also see that the success of the signal strength filtering can be improved significantly by installing sensors such that the pedestrians and noise are at different distances from the sensor and are the field of measurement is distinct from the surroundings however noisy it might be.

We then run the sequence numbers based clustering process on the rest of the probe requests to reduce the MAPE by almost 50 - 100\% on all the sensors except for location 3.
Location 3 is an outlier among all the other sensors since it is the only one with large amount of pedestrians very close to the sensor.
This may be causing the over filtering caused by the previous process.
We finally run the calibration process where we calculate the adjustment factors shown in Table \ref{errors_table}.
We the use them to adjust the counts to achieve a MAPE ranging from 10 - 50\%. We can see that the sensors with people moving right next to them tend to under-count with our methodology while sensors with seating next to them tend to over-count significantly. However, using the filtering process, we can reduce the error to almost 10\% closer to that of the ground truth.

>>>>

The methodology set out above is next implemented in five different Central
London locations at different times. Sensors are installed and data collected
for extended periods of time. We also carry out manual counting at these
locations across different times of the day. We then apply the methodologies
discussed earlier to arrive at estimated pedestrian footfall and compare them
with corresponding manual counts.  We finally evaluate the effectiveness of the
processes with the Mean Absolute Percentage Error (MAPE) at these locations and
report our findings.

@@

We start by looking at the distribution of the signal strength reported by the
probe requests across the locations. From the density plot shown in Figure
\ref{main_signal_strength} we can see that there is clear relation between the
distribution of the signal strength and the distance and complexity of the
source of noise. We can see that while location 5 shows clean difference between
low and high signal strengths, location 2 is almost normally distributed.
Intuitively we expect that location 2 and 4 must be harder to classify than
location 1 3 and 5.  We run the k-means clustering algorithm and filter out the
probe requests which are randomised and have signal strengths less than the
second break (threshold).  It is important to note that we are dealing with
relative thresholds of signal strengths which can vary with location and time of
the analysis.  We then aggregate then probe requests for every minute by
counting the number of Unique MAC addresses present in every minute. We also
remove devices that dwell around the sensor by removing the MAC addresses which
reappear from within the past hour.  The results of the first stage of filtering
process along with the thresholds are shown in Table \ref{errors_table}.
Confirming our intuition, we see that the location 2 has the most MAPE followed
by location 4 while rest of them have highly reduced MAPE.  It is significant
that this method alone reduces our margin of error by a 50 - 100\% from the raw
counts without any cleaning.  This makes the signal strength filtering quick and
ideal method for practical applications which doesn't require absolute numbers
such as creating large aggregated indexes to show long term trends.  We also see
that the success of the signal strength filtering can be improved significantly
by installing sensors such that the pedestrians and noise are at different
distances from the sensor and are the field of measurement is distinct from the
surroundings however noisy it might be.

We then run the sequence numbers based clustering process on the rest of the
probe requests to reduce the MAPE by almost 50 - 100\% on all the sensors except
for location 3.  Location 3 is an outlier among all the other sensors since it
is the only one with large amount of pedestrians very close to the sensor.  This
may be causing the over filtering caused by the previous process.  We finally
run the calibration process where we calculate the adjustment factors from the
ratio between the manual counts to the counts calculated for the sample period
as shown in Table \ref{errors_table}.  We the use them to adjust the counts to
achieve a MAPE ranging from 10 - 50\%. We can see that the sensors with people
moving right next to them tend to under-count with our methodology while sensors
with seating next to them tend to over-count significantly. However, using the
filtering process, we can reduce the error to almost 10\% closer to that of the
ground truth.

### 2018-11-20 paper/sections/methodology.tex
Tue 20 Nov 13:38:53 GMT 2018

Probe requests are a special type of management packets broadcast by Wi-Fi enabled devices as part of the various functions such as scanning for available APs, quick geolocation by triangulation based known APs, etc.
These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or model of the devices though there is some variation on the frequency and the information transmitted through them.
In some cases, such as Android devices, these are broadcast even when the Wi-Fi functionality has been turned off by the user.
Thus these signals can be used to reliably identify the presence of Wi-Fi enabled mobile devices.
Being a first step of connection initiated by the mobile device, these packets have information regarding the characteristics of the mobile device itself. Some of the key information we can infer from these requests are,
\begin{enumerate}
	\item \textbf{Media Access Control (MAC) address} which is an unique identifier for the wireless hardware of the mobile device,
	\item \textbf{Sequence number} of the request for the mobile device to keep track of the responses,
	\item \textbf{Time stamp} at which the request was received by the AP,
	\item Total \textbf{length} of the request in number of bits, and 
	\item The \textbf{strength of the signal} which transmitted the request.
\end{enumerate}
The MAC address is the primary unique identifier for the mobile device.
It has two parts, first part is the Organisation Unique Identifier (OUI) which gives information about the manufacturer of the device and the second is unique to the device. In modern devices, to protect the users privacy, the MAC address can also be randomised (hence non unique) and is marked as such. Though sequence number and length of the packet are not strictly unique, we hypothesize that we can use them to estimate the number unique devices.

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{images/datacollection_schematic.jpeg}
	\caption{Schematic diagram showing the process of collecting and storing probe requests using the sensor} \label{datacollection_schematic}
\end{figure}

The next step after collecting data was to estimate the footfall or pedestrian activity from them. We identified the following potential uncertainties which arise from our collection methodology.
\begin{enumerate}
	\item \textbf{Background noise} - since the extent to which Wi-Fi signals travel differs subject to various factors such as interference and humidity, it is close to impossible to restrict our data collection to a finite area of interest. This can lead to a significant background noise at certain locations. E.g. a phone shop or a bus stop located next to the study area can increase the number of probe requests received by the sensor.
	\item \textbf{MAC randomisation} - The mobile devices in recent years have been using randomised 'local' MAC addresses for probe requests to protect the users from being tracked. This makes it impossible to tell if the probe requests are being sent by the same mobile device which is being stationed next to the sensor. This along with the previous problem can further increase the magnitude of error by several fold.
	\item \textbf{Mobile ownership} - Since the rate of mobile ownership can vary widely across geography and demography, we cannot assume that every mobile device translates to one pedestrian footfall. In addition to this, there is a long term overall increase in mobile ownership which may lead to the number of probe requests collected overtime. 
\end{enumerate}
We propose the following internal and external validation methods to tackle each of these uncertainties.

\begin{figure}
	\centering
	\subfigure[Distribution of signal strengths (dBm) showing the filtering of background noise]{
		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_signalstrength.jpeg}}}\hspace{20pt}
	\subfigure[Clustering probe requests as nodes in a graph using increasing sequence numbers]{
	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_clustering.jpeg}}}
	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{methodology_schematic}
\end{figure}

\subsection{Filtering with Signal Strength}
One of the clues that we can use to estimate the distance between the mobile device and the sensor is the strength of the signal received by the sensor. 
The obvious approach here is to try and establish a relationship between the signal strength and distance first and use this to filter out the unwanted probe requests.
This approach was found not to be feasible since the decay of signal strength with distance is not always constant.
It varies with atmospheric conditions, presence of obstructions between the source and target, the nature of these obstructions and the strength (power level) of the source.
This severely limits our ability in establishing a simple conversion between reported signal strength and distance.
There is a need for a method which takes in to account these variables across various locations.

We hypothesise that in configurations where a specific source of background noise is at a constant distance, there must be a distinct pattern in the number of probe requests reporting signal strength corresponding to that distance.
For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported signal strength corresponding to the distance between the sensor and the phone shop irrespective of the local conditions as shown in Figure \ref{methodology_schematic}.
We could identify these breaks in the data using traditional one dimensional classification algorithms such as `jenks natural breaks', `k-means', `quantile' and `hierarchical clustering', etc.
Since we are only looking for the break in the data and not for absolute values, the methodology should apply for all the variations due micro site conditions thus reducing the overall noise in the collected data.

@@

The proposed algorithm creates a graph where the probe requests represented the nodes, and links are created between them based on the following rules: 
\begin{itemize}
	\item A link could go only forward in time. 
	\item A link could go from low to high sequence numbers.
	\item A link could exist between nodes with a maximum time difference of $\alpha$ - time threshold.
	\item A link could exist between nodes with a maximum sequence number difference of $\beta$ - sequence threshold.
	\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links in terms of both time and sequence number.
\end{itemize}
The nodes were then classified based on the unique connected component they belong to as shown in Figure \ref{methodology_schematic}.
This classification was assigned as the unique identifier for the anonymised probe requests in the place of MAC address.
Though the recycling of sequence number after 4000 leads to multiple classifications reported on single device, the magnitude of error is greatly reduced.

\subsection{Calibrating with Ground Truth}
Since mobile device ownership is an external uncertainty to our study and could arise from variety of spatio - temporal and demographic factors, we propose to solve this using external source of information.
We hypothesize that an adjustment factor could be arrived at for each location of data collection, comparing the sensor based counts and ground truth and it can be used to adjust the data reliably to reflect the ground truth in absolute numbers for the future.
This calibration can be carried over periodically and the frequency to improve the quality of the estimation.

>>>>

Probe requests are a special type of management packets broadcast by Wi-Fi
enabled devices as part of the various functions such as scanning for available
APs, quick geolocation by triangulation based known APs, etc. These are
broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or
model of the devices though there is some variation on the frequency and the
information transmitted through them. In some cases, such as Android devices,
these are broadcast even when the Wi-Fi functionality has been turned off by the
user. Thus these signals can be used to reliably identify the presence of Wi-Fi
enabled mobile devices. Being a first step of connection initiated by the mobile
device, these packets have information regarding the characteristics of the
mobile device itself. Some of the key information we can infer from these
requests are,

\begin{enumerate} 
\item \textbf{Media Access Control (MAC) address} which is an unique identifier
	for the wireless hardware of the mobile device, 
\item \textbf{Sequence number} of the request for the mobile device to keep
	track of the responses, 
\item \textbf{Time stamp} at which the request was received by the AP, 
\item Total \textbf{length} of the request in number of bits, and 
\item The \textbf{strength of the signal} which transmitted the request.
\end{enumerate}

The MAC address is the primary unique identifier for the mobile device. It has
two parts, first part is the Organisation Unique Identifier (OUI) which gives
information about the manufacturer of the device and the second is unique to the
device. In modern devices, to protect the users privacy, the MAC address can
also be randomised (hence non unique) and is marked as such. Though sequence
number and length of the packet are not strictly unique, we hypothesize that we
can use them to estimate the number unique devices as demonstrated by
\citep{vanhoef2016}.

@@

\begin{figure} 
\centering \includegraphics[width=\linewidth]
	{images/datacollection_schematic.jpeg}
\caption 
	{Schematic diagram showing the process of collecting and storing probe
	requests using the sensor}
\label{datacollection_schematic} 
\end{figure}

The next step after collecting data was to estimate the footfall or pedestrian
activity from them. We identified the following potential uncertainties which
arise from our collection methodology.

\begin{enumerate} 
\item 
\textbf{Background noise} - since the extent to which Wi-Fi signals travel
		differs subject to various factors such as interference and humidity, it
		is close to impossible to restrict our data collection to a finite area
		of interest. This can lead to a significant background noise at certain
		locations. E.g. a phone shop or a bus stop located next to the study
		area can increase the number of probe requests received by the sensor.
		It is important to note that this method may not work effectively on
		study locations with complex configurations such as the source of noise
		and the area of study being located at the same distance from the
		sensor. This aspect is explored in detail in the broader case study in
		the following sections.
\item 
\textbf{MAC randomisation} - The mobile devices in recent years have been using
		randomised 'local' MAC addresses for probe requests to protect the users
		from being tracked. This makes it impossible to tell if the probe
		requests are being sent by the same mobile device which is being
		stationed next to the sensor. This along with the previous problem can
		further increase the magnitude of error by several fold.
\item
\textbf{Mobile ownership} - Since the rate of mobile ownership can vary widely
		across geography and demography, we cannot assume that every mobile
		device translates to one pedestrian footfall. In addition to this, there
		is a long term overall increase in mobile ownership which may lead to
		the number of probe requests collected overtime.
\end{enumerate}

We propose the following internal and external validation methods to tackle each
of these uncertainties.

\begin{figure} 
\centering 
\subfigure [Distribution of signal strengths (dBm) showing the filtering of
		background noise] {\resizebox*{0.45\linewidth}{!}
		{\includegraphics{images/methodology_signalstrength.jpeg}}}
\hspace{20pt}
\subfigure [Clustering probe requests as nodes in a graph using increasing
		sequence numbers] {\resizebox*{0.45\linewidth}{!}
		{\includegraphics{images/methodology_clustering.jpeg}}}
\caption {Schematic diagrams explaining the methods for filtering by signal
	strength and clustering using sequence numbers}
\label{methodology_schematic}
\end{figure}

\subsection{Filtering with Signal Strength}

One of the clues that we can use to estimate the distance between the mobile
device and the sensor is the strength of the signal received by the sensor. The
obvious approach here is to try and establish a relationship between the signal
strength and distance first and use this to filter out the unwanted probe
requests. This approach was found not to be feasible since the decay of signal
strength with distance is not always constant. It varies with atmospheric
conditions, presence of obstructions between the source and target, the nature
of these obstructions and the strength (power level) of the source. This
severely limits our ability in establishing a simple conversion between reported
signal strength and distance. There is a need for a method which takes in to
account these variables across various locations.

We hypothesise that in configurations where a specific source of background
noise is at a constant distance, there must be a distinct pattern in the number
of probe requests reporting signal strength corresponding to that distance. For
example, if there is a phone shop next to our sensor where hundreds of phones
regularly send probe requests there should be a sharp rise of number of probe
requests with reported signal strength corresponding to the distance between the
sensor and the phone shop irrespective of the local conditions as shown in
Figure \ref{methodology_schematic}. We could identify these breaks in the data
using traditional one dimensional clustering algorithms such as `jenks natural
breaks', `k-means', `quantile' and `hierarchical clustering', etc. Since we are
only looking for the break in the data and not for absolute values, the
methodology should apply for all the variations due micro site conditions thus
reducing the overall noise in the collected data.

@@

The proposed algorithm creates a graph where the probe requests represented the
nodes, and links are created between them based on the following rules:

\begin{itemize} 
\item A link could go only forward in time. 
\item A link could go from low to high sequence numbers. 
\item A link could exist between nodes with a maximum time difference of
	$\alpha$ - time threshold. 
\item A link could exist between nodes with a maximum sequence number
	difference of $\beta$ - sequence threshold.
\item A node could have only one incoming link and one outgoing link, which is
	the shortest of all such possible links in terms of both time and sequence
	number.
\end{itemize}

The nodes were then assigned a unique id based on the unique connected component
they belong to as shown in Figure \ref{methodology_schematic}. This unique
identifier is used in the place of MAC address for aggregation for the
anonymised probe requests. Though the recycling of sequence number after 4096
leads to multiple unique ids being reported from a single device, the magnitude
of error is greatly reduced.

\subsection{Calibrating with Ground Truth}

Since mobile device ownership is an external uncertainty to our study and could
arise from variety of spatio - temporal and demographic factors, we propose to
solve this using external source of information. We hypothesize that an
adjustment factor could be arrived at for each location of data collection by
comparing the sensor based counts and ground truth and it can be used to adjust
the data reliably to reflect the ground truth in absolute numbers for the
future. This calibration can be carried out periodically and frequently to
improve the quality of the estimation over longer periods of time.
